Modeling the effect of aneuploidy on cancer evolution

### 2022-11-07 EvolutionaryRescue1.tex
Add files via upload

\begin{document}
\title{Evolutionary rescue}
\author[1]{Remus Stana}
\author[2]{‪Daniel Weissman}
\author[1]{Yoav Ram}
\affil[1]{School of Zoology, Tel Aviv University, Tel Aviv-Yafo, Israel}
\affil[2]{Department of Physics, Emory University, Atlanta, United States of America}
\maketitle
\begin{abstract}
Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction. A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments. We are interested in the mechanisms through which a population of cancer cells are able to adapt to chemotherapy, and in particular, the role played by chromosomal instability (aneuploidy). Cancer cells which have aneuploidy are hypothesized to have a higher fitness in an environment altered by anti-cancer drugs as they have incomplete pathways which drugs activate in order to kill the cells. Aneuploidy is highly prevalent in tumors and certain drugs which attempt to combat cancers through increasing chromosomal instability. As a result, the question we wish to answer is how aneuploidy impacts the fate of the population of cancer cells. We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive. Additionally, we will utilize large genomic datasets to asses the effects of aneuploidy on the probability of evolutionary rescue.
\end{abstract}

@@

\section{First case: $4\lambda_avp_m<\left(\lambda_a-\mu_a-v\right)^2$}
We assume that $\mu_w>\lambda_w$ and $\mu_a<\lambda_a$ and, as a result, we rewrite \eqref{survproba} and \eqref{survprobw} as:
\begin{align*}
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
\end{align*}
Making use of the Taylor series expansions:
\begin{align*}
\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}+\cdots\\
\left(1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_a-\mu_a-u-v\right)^2}+\cdots
\end{align*}
we obtain the following approximation for the survival probability of a population consisting of a single individual wildtype cell:
\begin{align}\label{survprobwinitial}
p_w&\approx\frac{vp_m+up_a}{\lambda_a-\mu_a-u-v}\\
\nonumber
&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw}
&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right],
\end{align}
where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
\begin{equation}\label{notationalconv}
\Delta_i=\lambda_i-\mu_i,
\end{equation}
we write \eqref{survprobw} as
\begin{equation}\label{survprobwapprox}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right],
\end{align}
which we plot in Figure \ref{SurvPlotNData} as a function of $N$ and in Figure \ref{SurvPlot} as a function of $\lambda_w$.

We want to improve the accuracy of our approximation by taking into consideration the second term of the Taylor series expansion:
\begin{align*}
\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
\end{align*}
which gives us the following approximation for $p_a$:
\begin{align}
p_a=\frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3}
\end{align}
From which we deduce that:
\begin{align}\nonumber
p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw}
&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right].
\end{align}
Using the notations described in \eqref{notationalconv} we write the above equation as:
\begin{equation}\label{survprobwapproxcorrected}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right].
\end{align}
\section{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
If we assume that  $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$ then we write:
\begin{equation}
p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a}
\end{equation}
and using the following Taylor series expansion:
\begin{align*}
\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots
\end{align*}
we obtain:
\begin{align*}
p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
&\approx\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}
\end{align*}
In the last line of the above we have used the following inequalities:
\begin{align*}
\lambda_a-\mu_a,\,v,\,\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}\ll1.
\end{align*}
As a result, we have from \eqref{survprobwinitial} the probability of rescue of a population starting from one wildtype individual:
\begin{align*}
p_w&\approx\frac{1}{\lambda_a-\mu_a-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right]\\
&=\sqrt{\frac{\lambda_m-\mu_m}{\lambda_m}}\frac{\sqrt{v}}{\lambda_a-\mu_a-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}}\right]\\
&=\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_a-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right],
\end{align*}
where in the last line we have used the notations defined in \eqref{notationalconv}.

@@

\bibliographystyle{unsrt}
\bibliography{green2021}

@@

\section{Survival probability of a mutant lineage}\label{AppendixSurvLin}
\begin{align*}
p_{i+j,i}\left(\Delta t\right)=\left\{
  \begin{array}{@{}ll@{}}
  \mu i\Delta t+o\left(\Delta t\right),\quad j=-1 \\
  \lambda i\Delta t+o\left(\Delta t\right),\quad j=1 \\
    1-\left(\lambda+\mu\right) i\Delta t+o\left(\Delta t\right),\quad j=0 \\
   o\left(\Delta t\right),\quad j\neq-1,0,1.
  \end{array}\right.
\end{align*}

\begin{align*}
\frac{dp_i\left(t\right)}{dt}&=\lambda\left(i-1\right)p_{i-1}\left(t\right)+\mu\left(i+1\right)p_{i+1}\left(t\right)-\left(\lambda+\mu\right)ip_i\left(t\right)\\
\frac{dp_0\left(t\right)}{dt}&=\mu p_1\left(t\right)
\end{align*}

\begin{align*}
\mathcal{P}\left(z,t\right)=\left\{
  \begin{array}{@{}ll@{}}
  \left(\frac{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\mu\left(z-1\right)}{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\lambda\left(z-1\right)}\right)^N,\quad \text{if }\lambda\neq\mu \\
  \left(\frac{1-\left(\lambda t-1\right)\left(z-1\right)}{1-\lambda t\left(z-1\right)}\right)^N,\quad \text{if }\lambda=\mu
  \end{array}\right.
\end{align*}

\begin{equation*}
\mathcal{P}\left(z,t\right)=\sum_{i=0}^\infty p_i\left(t\right)z^i \quad\text{and}\quad p_i\left(t\right)=\left.\frac{1}{i!}\frac{\partial^i\mathcal{P}}{\partial z^i}\right|_{z=0}
\end{equation*}

\begin{align*}
p_0\left(t\right)=\left\{
  \begin{array}{@{}ll@{}}
  \left(\frac{\mu-\mu\e^{\left(\mu-\lambda\right)t}}{\lambda-\mu\e^{\left(\mu-\lambda\right)t}}\right)^N,\quad \text{if }\lambda\neq\mu \\
  \left(\frac{\lambda t}{1+\lambda t}\right)^N,\quad \text{if }\lambda=\mu
  \end{array}\right.
\end{align*}

\begin{align*}
p_0\left(\infty\right)=\lim_{t\rightarrow\infty}p_0\left(t\right)=\left\{
  \begin{array}{@{}ll@{}}
  1,\quad \text{if }\lambda\leq\mu \\
  \left(\frac{\mu}{\lambda}\right)^N,\quad \text{if }\lambda>\mu
  \end{array}\right.
\end{align*}
\end{document}

>>>>

\begin{document}
\title{Modeling the effect of aneuploidy on cancer evolution}
\author[1]{Remus Stana}
\author[2]{‪Daniel Weissman}
\author[1]{Yoav Ram}
\affil[1]{School of Zoology, Tel Aviv University, Tel Aviv-Yafo, Israel}
\affil[2]{Department of Physics, Emory University, Atlanta, United States of America}
\maketitle
\begin{abstract}
Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction. A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments. We are interested in the mechanisms through which a population of cancer cells are able to adapt to chemotherapy, and in particular, the role played by chromosomal instability (aneuploidy). Cancer cells which have aneuploidy are hypothesized to have a higher fitness in an environment altered by anti-cancer drugs as they have incomplete pathways which drugs activate in order to kill the cells. Aneuploidy is highly prevalent in tumors and certain drugs which attempt to combat cancers through increasing chromosomal instability. As a result, the question we wish to answer is how aneuploidy impacts the fate of the population of cancer cells. We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive. Additionally, we will utilize large genomic datasets to asses the effects of aneuploidy on the probability of evolutionary rescue.
\end{abstract}

@@

\section{First case: $4\lambda_avp_m<\left(\lambda_a-\mu_a-v\right)^2$}
\subsection{$\lambda_a>\mu_a$ and $\lambda_w<\mu_w$}
We assume that $\lambda_a>\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
\begin{align*}
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
\end{align*}
Making use of the Taylor series expansions:
\begin{align*}
\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}+\cdots\\
\left(1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}+\cdots
\end{align*}
we obtain the following approximation for the survival probability of a population consisting of a single individual wildtype cell:
\begin{align}\label{survprobwinitial}
p_w&\approx\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
\nonumber
&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right],
\end{align}
where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
\begin{equation}\label{notationalconv}
\Delta_i=\lambda_i-\mu_i,
\end{equation}
we write \eqref{survprobw} as
\begin{equation}\label{survprobwapprox}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right],
\end{align}
which we plot in Figure \ref{SurvPlotNData} as a function of $N$ and in Figure \ref{SurvPlot} as a function of $\lambda_w$.

We want to improve the accuracy of our approximation by taking into consideration the second term of the Taylor series expansion:
\begin{align*}
\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
\end{align*}
which gives us the following approximation for $p_a$:
\begin{align}
p_a=\frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3}
\end{align}
From which we deduce that:
\begin{align}\nonumber
p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right].
\end{align}
Using the notations described in \eqref{notationalconv} we write the above equation as:
\begin{equation}\label{survprobwapproxcorrected}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right].
\end{align}
\subsection{$\lambda_a<\mu_a$ and $\lambda_w<\mu_w$}
We assume that $\lambda_a<\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
\begin{align*}
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
\end{align*}
\begin{align}\label{survprobwinitial}
p_w&\approx\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
\nonumber
&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
&\approx\frac{v}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u\left(\lambda_m-\mu_m\right)}{\left(\lambda_a-\mu_a\right)}-\left(\lambda_m-\mu_m\right)\right],
\end{align}
where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
\begin{equation}\label{notationalconv}
\Delta_i=\lambda_i-\mu_i,
\end{equation}
we write \eqref{survprobw} as
\begin{equation}\label{survprobwapprox}
p_w=\frac{v}{\lambda_m\Delta_w}\left(\frac{u\Delta_m}{\Delta_a}-\Delta_m\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{vN}{\lambda_m\Delta_w}\left(\Delta_m-\frac{u\Delta_m}{\Delta_a}\right)\right],
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
If we assume that  $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$ then we write:
\begin{equation}
p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a}
\end{equation}
and using the following Taylor series expansion:
\begin{align*}
\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots
\end{align*}
we obtain:
\begin{align*}
p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
&\approx\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}
\end{align*}
In the last line of the above we have used the following inequalities:
\begin{align*}
\lambda_a-\mu_a,\,v,\,\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}\ll1.
\end{align*}
As a result, we have from \eqref{survprobwinitial} the probability of rescue of a population starting from one wildtype individual:
\begin{align*}
p_w&\approx\frac{1}{\lambda_a-\mu_a-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right]\\
&=\sqrt{\frac{\lambda_m-\mu_m}{\lambda_m}}\frac{\sqrt{v}}{\lambda_a-\mu_a-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}}\right]\\
&=\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_a-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right],
\end{align*}
where in the last line we have used the notations defined in \eqref{notationalconv}.

@@

\bibliographystyle{unsrt}
\bibliography{evo2022}

@@

\section{Survival probability of a mutant lineage}\label{AppendixSurvLin}
The infinitesimal transition probabilities for the simple birth and death process:
\begin{align*}
p_{i+j,i}\left(\Delta t\right)=\left\{
  \begin{array}{@{}ll@{}}
  \mu i\Delta t+o\left(\Delta t\right),\quad j=-1 \\
  \lambda i\Delta t+o\left(\Delta t\right),\quad j=1 \\
    1-\left(\lambda+\mu\right) i\Delta t+o\left(\Delta t\right),\quad j=0 \\
   o\left(\Delta t\right),\quad j\neq-1,0,1.
  \end{array}\right.
\end{align*}
The forward Kolmogorov differential equations are:
\begin{subequations}
\label{KolmeqSurv}
\begin{flalign}
\frac{dp_i\left(t\right)}{dt}&=\lambda\left(i-1\right)p_{i-1}\left(t\right)+\mu\left(i+1\right)p_{i+1}\left(t\right)-\left(\lambda+\mu\right)ip_i\left(t\right),\\
\frac{dp_0\left(t\right)}{dt}&=\mu p_1\left(t\right),
\end{flalign}
\end{subequations}
for $i=1,2,\dots$ with initial conditions $p_i\left(0\right)=\delta_{iN}$.

The probability generating function is defined as:
\begin{equation*}
\mathcal{P}\left(z,t\right)=\sum_{i=0}^\infty p_i\left(t\right)z^i
\end{equation*}
which we obtain by multiplying \eqref{KolmeqSurv} by $z^i$ and summing over $i$:
\begin{align*}
\mathcal{P}\left(z,t\right)=\left\{
  \begin{array}{@{}ll@{}}
  \left(\frac{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\mu\left(z-1\right)}{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\lambda\left(z-1\right)}\right)^N,\quad \text{if }\lambda\neq\mu, \\
  \left(\frac{1-\left(\lambda t-1\right)\left(z-1\right)}{1-\lambda t\left(z-1\right)}\right)^N,\quad \text{if }\lambda=\mu.
  \end{array}\right.
\end{align*}
The probability $p_i$ can be obtained from the probability generating function as:
\begin{equation*}
p_i\left(t\right)=\left.\frac{1}{i!}\frac{\partial^i\mathcal{P}}{\partial z^i}\right|_{z=0},
\end{equation*}
and the probability of extinction is given by:
\begin{align*}
p_0\left(t\right)=\left\{
  \begin{array}{@{}ll@{}}
  \left(\frac{\mu-\mu\e^{\left(\mu-\lambda\right)t}}{\lambda-\mu\e^{\left(\mu-\lambda\right)t}}\right)^N,\quad \text{if }\lambda\neq\mu, \\
  \left(\frac{\lambda t}{1+\lambda t}\right)^N,\quad \text{if }\lambda=\mu.
  \end{array}\right.
\end{align*}
When $t\rightarrow\infty$ the extinction probability has the following expression:
\begin{align*}
p_0\left(\infty\right)=\lim_{t\rightarrow\infty}p_0\left(t\right)=\left\{
  \begin{array}{@{}ll@{}}
  1,\quad \text{if }\lambda\leq\mu, \\
  \left(\frac{\mu}{\lambda}\right)^N,\quad \text{if }\lambda>\mu.
  \end{array}\right.
\end{align*}
\end{document}
